Diffusion of Innovation in the Gentoo Portage Ecosystem

### 2014-08-21 Paper/paper.tex
Changed the following 1. Title 2. Minor changes to Intro 3. More changes to the Literature Background section 4. Added more references

\usepackage{amsmath}
\usepackage[round]{natbib}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{microtype} % Finetunes spacing, load after fonts
\usepackage{siunitx}
\usepackage{alltt}
\usepackage{fancyvrb}
\usepackage{wrapfig}
\usepackage[colorlinks=false, pdfborder={0 0 0}]{hyperref} % Should be loaded last
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{subfigure}
\newcommand{\nd}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\tn}{\newline}

@@

\title{Gentoo Package Dependencies over Time}
\subtitle{Subtitle}
\dedication{}
\author{Remco Bloemen \and Chintan Amrit}
\institute{University of Twente}

@@

\tableofcontents

@@

Open source software has had a large impact on various aspects of software development and has been the source of  innovations in software technology, development process and even in software development business models \citep{ebert07}. Hence identifying innovative software projects as well as analysing their adoption is essential to arrive at an understanding of the diffusion of innovation of Open source projects. However, literature on adoption of open source projects is at a preliminary stage, largely focussed on adoption of popular open source projects by organisations and end users \citep{aksulu10}.There is especially a dearth of research regarding the adoption of open source libraries (or even subsystems \cite{maccormack06}) that are the building blocks of large software projects \citep{szyperski04}. There is also a corresponding dearth of knowledge on the methods one can use to test the extent of adoption of software libraries over a period of time.

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the lifecycle of the adoption. Our method is borrowed from the field of marketing research \citep{bass69}. In particular, we study  software innovation of the Gentoo Portage libraries from the adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass Diffusion model \citep{bass69}. We then test the fit of the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) to the Bass Diffusion Model. In all cases we find that the Bass Diffusion model explains the adoption of the libraries - with an $\overline{R}^2 > 98\%$. This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass Diffusion model. This finding, if found generalisable to most large open source libraries can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

% (Adoption reflects innovation 'better' than changes within packages. - Still cannot find a relevant reference!)%

@@

\subsection{ Open Source Software Ecosystems}
No software project stands entirely on its own. Software is usually developed by taking one or more existing libraries of components and combining those components in ways to create new products. Take for example a simple chat application. The chat application uses a different library for user interface development that provides components such as a window a text entry field and a button (that is labeled "send message" by the chat application). This user interface library in its turn uses a graphics library to draw the lines, rectangles and text necessary for the fields and buttons. The graphics library uses a library to read font files and uses the fonts to turn text into pictures that can be displayed on the screen. The graphics library then sends the contents of the window to the window manager, which in turns uses a graphics card driver to instruct the hardware. The chat application uses a networking library to provide it with the basic components for internet communication and uses a file library to store the users settings. The same file library is also used by the font library to read font files.

In \citet{crowston08} published a comprehensive overview of academical research on open source software development. Of the 184 articles they cite, the vast majority of articles are case studies or surveys, with 4\% of the articles describing the development of empirical instruments and/or measurements. Also, most articles look at the level of a particular group or project, with 4\% looking at the societal level of interacting projects. Crowston et al. (2008) find no articles that develop instruments or measurements on the level of OSS packages, which this paper aims to do \citep{crowston08} .

\citet{haefliger08} study code re-use on six open source projects, they conclude that there is extensive code re-use in open source software. The article proceeds by identifying the process of code re-use, such as the drivers for re-using and the tools used to find relevant code. The study by Haefliger et al. looks at a given project and how it  re-uses existing components \citep{haefliger08}, whereas this paper looks at a given project and how it is being re-used by other projects. Additionally, we performed our analysis using a large set of automatically collected and processed empirical data.

@@

@@

Now that a data source is selected, it is time to extract the required information and process the data into a form that allows easy calculations. The major steps are collecting the raw data, parsing this into a simpler format and producing the final dependency graph from this simpler form. Some post-processing can then be done on the dependency graph. In the process, the dataset will shrink from thirteen gigabytes taking more than a week to collect, to thirty megabytes that can be processed in four seconds.

@@

German et al. The Evolution of the R Software Ecosystem

@@

To model the process of innovation diffusion \cite{bass69} introduces two processes that propagate an innovation. The first processes is involves individuals that decide to use an innovation based on their perception of its merits, without looking at the experiences of others. The second process involves the word-of-mouth effect or the bandwagon effect, individuals adopt the innovation solely because they hear of the experiences of previous adopters. Of course in reality, everyone will be somewhere in between these two extreme types, but for the sake of modelling it suffices to consider the relative abundance of both types.

It should be noted that \cite{bass69} and all later authors, use confusing terms to describe the two types of adopters. The first type are called "innovators", not to be confused with those actually inventing the innovation and the second type are called "imitators", not to be confused with those developing imitating offerings. If one remembers that the model concerns the demand side of the market and not the supply side than it will all be clear.

@@

\cite{bass69} solves this ordinary differential equation, which results in the following function for $F$:

@@

Equation~\eqref{eq:bass-model} incorporates the two additions and can be readily applied to empirical data. In \cite{mahajan95} and many other empirical studies this happens in the differential form, since only absolute sales figures are available and not absolute user figures. The dependency graph method presented later allows one to obtain absolute usage number, so the differential form is not further used.

@@

% Package provides
Using all the observations and choices made above, a tool was developed to extract dependency graphs from all the simplified database snapshots. The shear scale of the resulting dataset, $1.3$ million packages and $6.9$ million dependency relations, required a solution to store efficiently. Therefore, a compressed format was developed that only stores the changes between the historical snapshots instead of storing whole snapshots. The extraction process took three hours and resulted in a $29$ megabyte file. Reading this file into memory resident data structures and performing simple queries takes about $4$ seconds. The dataset is now in a usable form.

\subsection{Final Remarks}

@@

The package dependency graph contains empirical data to test extensions of the Bass diffusion model - extended with discarders. The Bass model and the present analysis is formulated in terms of absolute number of users, but in most applications only sales figures are available. The amount of sales is the first derivative of the Bass model, hence the model is usually applied in its derivative form \citep{mahajan95}. As a consequence the model only considers adopters, but does not consider \emph{discarders}. In the \verb|xulrunner| example, we see the package being discarded from 2011 onwards, providing insights in the discarding mechanism. The next step would be to collect more examples of packages being discard, look at their patterns of demise and develop a model of discarding to supplement the Bass model of adoption. One model could for example be the inverse of a Bass curve, this makes sense when the market share of the original package is taken over by a new package. The unique feature of dependency graph analysis to give absolute user numbers facilitates this. Such an analysis can also help in predicting the behaviour of certain projects - hence forewarning the particular project stakeholders. Hence, we can typically supplement the existing analysis of open source project health \citep{daniel13},\citep{amrit10a},\citep{piggot13}. As the extent of adoption of an open source project is an indication of how popular the project is and in-turn an indication of the health of the project as assessed by the open source community.

@@

\bibliographystyle{plainnat}
\bibliography{paper}
\end{document}

>>>>

\usepackage{amsmath}
\usepackage[round]{natbib}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{microtype} % Finetunes spacing, load after fonts
\usepackage{siunitx}
\usepackage{alltt}
\usepackage{fancyvrb}
\usepackage{wrapfig}
\usepackage[colorlinks=false, pdfborder={0 0 0}]{hyperref} % Should be loaded last
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{subfigure}
\usepackage{cite}
\newcommand{\nd}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\tn}{\newline}

@@

\title{Diffusion of Innovation in the Gentoo Portage Ecosystem}
\subtitle{Do we need a subtitle?} 
\dedication{}
\author{Remco Bloemen \and Chintan Amrit}
\institute{University of Twente}

@@

%\tableofcontents

@@

Open source software has had a large impact on various aspects of software development and has been the source of  innovations in software technology, development process and even in software development business models \citep{ebert07}. Hence identifying innovative software projects as well as analysing their adoption is essential to arrive at an understanding of the diffusion of innovation of Open source projects. However, literature on adoption of open source projects is at a preliminary stage, largely focussed on adoption of popular open source projects by organisations and end users \citep{aksulu10}.There is especially a dearth of research regarding the adoption of open source libraries (or even subsystems \citep{maccormack06}) that are the building blocks of large software projects \citep{szyperski04}. There is also a corresponding dearth of knowledge on the methods one can use to test the extent of adoption of software libraries over a period of time.

In this paper we not only analyse the adoption of open source libraries by other projects, but also use a method to test the diffusion of innovation of a software project - or the lifecycle of a project's adoption. Our method is borrowed from the field of marketing research \citep{bass69}. In particular, we study  software innovation of the libraries in the Gentoo Portage  ecosystem from an adoption perspective. We first download the Gentoo Portage database for a time line of one month over a period of around 13 years (from 1 January to around mid 2012). We then calculate the dependencies among the libraries in each version of the database. Then, we calculate the change in the dependency graph of some well known libraries and check if this adoption of the library fits the Bass Diffusion model \citep{bass69}. We then test the fit of the adoption of 10 randomly selected libraries (we discard and replace those libraries whose adoption trends contain a lot of noise due to various reasons) to the Bass Diffusion Model. In all cases we find that the Bass Diffusion model explains the adoption of the libraries - with an $\overline{R}^2 > 98\%$. This finding shows that the adoption of open source software libraries (in the Gentoo Portage database) does not differ from the adoption of other kinds of innovations as predicted by the theory of adoption of innovation, and in particular the Bass Diffusion model. This finding, if found generalisable to most large open source libraries, can help in predicting their adoption over their lifetime. This can hence provide guidance to package managers and users of open source libraries.

% (Adoption reflects innovation 'better' than changes within packages. - Still cannot find a relevant reference!)

@@

\subsection{ Open Source Software Ecosystems}
No software project stands entirely on its own. Software is usually developed by taking one or more existing libraries of components and combining those components in ways to create new products. Take for example a simple chat application. The chat application uses a different library for user interface development that provides components such as a window a text entry field and a button (that is labeled "send message" by the chat application). This user interface library in turn uses a graphics library to draw the lines, rectangles and text necessary for the fields and buttons. The graphics library uses a library to read font files and uses the fonts to turn text into pictures that can be displayed on the screen. The graphics library then sends the contents of the window to the window manager, which in turn uses a graphics card driver to instruct the hardware, and so on.  This collection of software projects that interact under the umbrella of a common software platform is variously been called as a software "ecosystem" \citep{manikas13}. In this paper we adopt Lungu's short and elegant description of an ecosystem as "\textit{a collection of software projects which are developed and co-evolve in the same environment}" \citep{lungu10}.

The research on software ecosystems has primarily focussed on their business within a shared market or the relationships among the different businesses and hence deals with web and mobile platforms \citep{german13}.  Research on open source software ecosystems has focussed on the debian linux distribution \citep{gonzalez09, german07}, the linux kernel \citep{godfrey00}, the eclipse platform \citep{wermelinger08} and recently GNU R \citep{german13}. They analyse the evolution of the ecosystems as growth in terms of number of packages and the dependencies among the packages and not in terms of adoption of innovation as we do in this paper.

In \citet{crowston08} published a comprehensive overview of academical research on open source software development. Of the 184 articles they cite, the vast majority of articles are case studies or surveys, with 4\% of the articles describing the development of empirical instruments and/or measurements. Also, most articles look at the level of a particular group or project, with 4\% looking at the societal level of interacting projects. Crowston et al. (2008) find no articles that develop instruments or measurements on the level of OSS packages \citep{crowston08}, which we aim to do in this paper.

\citet{haefliger08} study code re-use on six open source projects, they conclude that there is extensive code re-use in open source software. The article proceeds by identifying the process of code re-use, such as the drivers for re-using and the tools used to find relevant code. The study by Haefliger et al. (2008) looks at a given project and how it  re-uses existing components \citep{haefliger08}, whereas in this paper we analyse a given project and how it is being re-used by other projects. Additionally, we perform our analysis using a large set of automatically collected and processed empirical data.

@@

\citet{lungu10} present multiple methods to automatically recover dependencies between software projects in an ecosystem. They arrive at a representation of the ecosystem that they call \textit{Ecco} that contains sufficient data to extract the dependencies of the ecosystem \citep{lungu10}.

\citet{german13} study the ecosystem of the GNU R project and determine the differences between the user-contributed packages and the core packages. They study the code characteristics, code evolution, dependency structures and the community surrounding the core and user-contributed packages. They find some significant significant differences in many of these aspects, and in particular they find that though user-contributed packages have been growing at a fast rate than the core packages, they remain stable in size over time \citep{german13}. Of particular interest to this paper, they examine the dependency structure of the R packages by analysing the DESCRIPTION file of each packages that describes the list of packages required by the particular package for proper installation. They then build an inter-dependency graph based on these package installation dependencies \citep{german13}. Their main findings related to this aspect is that there are consistently small number of dependencies for each package and that the peripheral packages have fewer dependencies \citep{german13}.

@@

Now that a data source is selected, it is time to extract the required information and process the data into a form that allows easy calculations. The major steps are collecting the raw data, parsing this into a simpler format and producing the final dependency graph from this simpler form. Some post-processing can then be done on the dependency graph. In the process, the dataset shrunk from thirteen gigabytes (taking more than a week to collect), to thirty megabytes that could be processed in four seconds.

@@

@@

To model the process of innovation diffusion \citet{bass69} introduces two processes that propagate an innovation. The first processes is involves individuals that decide to use an innovation based on their perception of its merits, without looking at the experiences of others. The second process involves the word-of-mouth effect or the bandwagon effect, individuals adopt the innovation solely because they hear of the experiences of previous adopters. Of course in reality, everyone will be somewhere in between these two extreme types, but for the sake of modelling it suffices to consider the relative abundance of both types.

It should be noted that \citet{bass69} and all later authors, use confusing terms to describe the two types of adopters. The first type are called "innovators", not to be confused with those actually inventing the innovation and the second type are called "imitators", not to be confused with those developing imitating offerings. If one remembers that the model concerns the demand side of the market and not the supply side than it will all be clear.

@@

\citet{bass69} solves this ordinary differential equation, which results in the following function for $F$:

@@

Equation~\eqref{eq:bass-model} incorporates the two additions and can be readily applied to empirical data. In \citep{mahajan95} and many other empirical studies this happens in the differential form, since only absolute sales figures are available and not absolute user figures. The dependency graph method presented later allows one to obtain absolute usage number, so the differential form is not further used.

@@

% Package provides
Using all the observations and choices made above, a tool was developed to extract dependency graphs from all the simplified database snapshots. The shear scale of the resulting dataset, $1.3$ million packages and $6.9$ million dependency relations, required a solution to store efficiently. Therefore, a compressed format was developed that only stores the changes between the historical snapshots instead of storing whole snapshots. The extraction process took three hours and resulted in a $29$ megabyte file. Reading this file into memory resident data structures and performing simple queries takes about $4$ seconds. The dataset is now in a usable form. The dataset and the details of the data structure can be seen in Appendix 1.

@@

The package dependency graph contains empirical data to test extensions of the Bass diffusion model - extended with discarders. The Bass model and the present analysis is formulated in terms of absolute number of users, but in most applications only sales figures are available. The amount of sales is the first derivative of the Bass model, hence the model is usually applied in its derivative form \citep{mahajan95}. As a consequence the model only considers adopters, but does not consider \emph{discarders}. In the \verb|xulrunner| example, we see the package being discarded from 2011 onwards, providing insights in the discarding mechanism. The next step would be to collect more examples of packages being discard, look at their patterns of demise and develop a model of discarding to supplement the Bass model of adoption. One model could for example be the inverse of a Bass curve, this makes sense when the market share of the original package is taken over by a new package. The unique feature of dependency graph analysis to give absolute user numbers facilitates this. Such an analysis can also help in predicting the behaviour of certain projects - hence forewarning the particular project stakeholders. As the extent of adoption of an open source project is an indication of how popular the project is and in-turn an indication of the health of the project as assessed by the open source community, we can typically supplement the existing analysis of open source project health \citep{daniel13,amrit10a,piggot13}.

@@

\bibliographystyle{plainnat}
\bibliography{paper}

\appendix
\section{Dataset}

\end{document}
